---
title: "Evans_Version"
format: html
Author: Evan Whitfield (For this file)
---

# Task 1 - Querying an API and a Tidy-Style Function

## Number 1

```{r}
library(httr)
response <- GET("https://newsapi.org/v2/everything?q=Iran&from=2026-05-20&language=en&pageSize=50&apiKey=173ec253cb06408abbf28061323e5fb4")

str(response, max.level = 1)
```

Look at the `content`.

```{r}
library(jsonlite)
library(tibble)
parsed_response <- fromJSON(rawToChar(response$content))
str(parsed_response, max.level = 1)
```

Pull a single article.

```{r}
trump_article <- parsed_response$articles[4, ]
str(trump_article, max.level = 1)
```

Ehh. Let's print it out better.

```{r}
print(trump_article)
```
```{r}
trump_article$content
```


## Function Making

```{r}
article_finder <- function(subject, 
                           date = format(Sys.Date() - 30, "%Y-%m-%d"), 
                           key = "173ec253cb06408abbf28061323e5fb4") {
  URL <- paste0("https://newsapi.org/v2/everything?q=",
                 subject,
                 "&from=",
                 date,
                 "&language=en&pageSize=50&apiKey=",
                 key)
  response <- httr::GET(URL)
  
  parsed_response <- fromJSON(rawToChar(response$content))
}
```

```{r}
football_articles <- article_finder("Football")
str(football_articles, max.level = 2)
```

```{r}
Liverpool_articles <- article_finder("Liverpool")
str(Liverpool_articles, max.level = 2)
```

```{r}
cat(Liverpool_articles$articles[1,]$content)
```

# Task 2 - Summarizing Roller Coaster Data

## Read in Data
```{r}
#| warning: False
library(tidyverse)
coaster_data <- read_csv("../HW4DataFiles/635USrollercoasters.csv")
```
## Basic EDA

Initial Data Look
```{r}
str(coaster_data)
```
It does not look like any of the variables (num or chr) need to be switched. We will want to switch some factors.
```{r}
coaster_data
```

Removing Latitude and Longitude 
```{r}
coaster_data <- coaster_data |>
  select(-"Latitude", -"Longitude")

psych::describe(coaster_data)
```

Now let's look and see the rate of missing values.
```{r}
colSums(is.na(coaster_data))
```

## Categorical Variables

## Numeric Variables (and across groups)




