---
title: "HW #4"
author: "Ryan Friedman, Evan Whitfield, Ralph Church"
format: html
editor: visual
---

```{r}
#| include: false
# load packages that may be used
library(tidyverse)
library(readxl)
library(DBI)
library(knitr)
library(httr)
library(jsonlite)
library(psych)
library(gt)
```

## Task 1 - Querying an API and a Tidy-Style Function

1.  Use GET() from the httr package to return information about a topic that you are interested in that has been in the news lately (store the result as an R object).

```{r}
RFKey <- "27554db1fa15488a995b7935d17bdf94" # R.Friedman API key for newsapi.org

APIData <- GET(paste0("https://newsapi.org/v2/everything?q=SpaceX&from=2026-06-01&language=en&pageSize=50&apiKey=", RFKey)) # Gets data on topic "SpaceX" from 6/1/2026 onward.
```

2.  Parse what is returned and find your way to the data frame that has the actual article information in it.

```{r}
parsed <- fromJSON(rawToChar(APIData$content))
parsed$articles
```

3.  Write a function that allows the user to easily query NewsAPI. The inputs are the title/subject to search for, a time period to search from (i.e. from that time until the present), and an API key.

```{r}
NewsAPIQuery <- function (subject,TimePeriod, NewsAPIKey){
  NewsURL <- paste0("https://newsapi.org/v2/everything?q=", subject,"&from=", TimePeriod, "&language=en&pageSize=50&apiKey=", NewsAPIKey) # creates the URL from the arguments
  NewsData <- GET(NewsURL) # gets the API data
  NewsParsed <- fromJSON(rawToChar(NewsData$content)) # convert content of API data into readable test
  NewsParsed$articles # returns articles dataframe
}
```

Testing function with search of "Disneyland" from "2026-06-08".

```{r}
NewsAPIQuery("Disneyland", "2026-06-08", RFKey)
```

Testing function with search of "Tokyo" from "2026-06-04".

```{r}
NewsAPIQuery("Tokyo", "2026-06-04", RFKey)
```

## Task 2 - Summarizing Roller Coaster Data

- Read in the data. Confirm variable data types and identify factors

```{r}
CoasterDataRaw <-read_csv("HW4DataFiles/635USrollercoasters.csv")
CoasterDataRaw
```

- Delete Latitude, Longitude, and Boundary variables (not needed for the assignment)
- Fix variable names
- Variables to be converted to factors: CensusRegion, AgeGroup, Scale, Type, Design, InversionsYN

```{r}
CoasterDataCleaned <- CoasterDataRaw |>
  select(-Latitude, -Longitude, -Boundary) # Remove undesired variables

colnames(CoasterDataCleaned) = c("Coaster", # Set variable names
                                 "Park",
                                 "City",
                                 "State",
                                 "CensusRegion",
                                 "YearOpened",
                                 "AgeGroup",
                                 "TopSpeed",
                                 "MaxHeight",
                                 "Scale",
                                 "Type",
                                 "Design",
                                 "Drop",
                                 "Length",
                                 "Duration",
                                 "InversionsYN",
                                 "NumInversions",
                                 "Make")  

CoasterDataCleaned <- CoasterDataCleaned |>
  mutate(CensusRegion = factor(CensusRegion), # Convert select variables to factors
         AgeGroup = factor(AgeGroup),
         Scale = factor(Scale),
         Type = factor(Type),
         Design = factor(Design),
         InversionsYN = factor(InversionsYN))

str(CoasterDataCleaned) # Confirm remaining variables and types
```

- Check for NA values in the data

```{r}
colSums(is.na(CoasterDataCleaned))
```

Note: There are no NA values for the factor variables.

### Categorical Variables

- Use the table() function to create a one-way contingency table for Design

```{r}
table(CoasterDataCleaned$Design)
```

Note: The most common design is 'Sit Down', with 560 out of the total 635 roller coasters in the data set.

- Use the table() function to create a two-way contingency table for Type by InversionsYN

```{r}
table(CoasterDataCleaned$Type, CoasterDataCleaned$InversionsYN)
```

Note: Wood roller coasters with inversions are quite rare. The data set only identifies 4 out of the total 635.

- Use the table() function to create a three-way contingency table for AgeGroup by Scale by CensusRegion

```{r}
table(CoasterDataCleaned$AgeGroup, CoasterDataCleaned$Scale, CoasterDataCleaned$CensusRegion)
```

Note: Across the three factors of age, scale, and region, the most common combination is Newest/Family/South, with 55 of 635 coasters

- Use the table() and filter() functions to create a conditional two-way table for Scale by CensusRegion among Newest (AgeGroup) roller coasters

```{r}
FilteredData <- CoasterDataCleaned |>
  filter(AgeGroup == "3:Newest") 

table(FilteredData$Scale, FilteredData$CensusRegion)
```

- Recreate the conditional two-way table above using a subsetted three-way table

```{r}
table(CoasterDataCleaned$Scale, CoasterDataCleaned$CensusRegion, CoasterDataCleaned$AgeGroup)[ , , 3]
```

- Create a two-way contingency table for Type by Design using group_by() and summarize() from dplyr. Then use pivot_wider() to make the result look more like the output from table().

```{r}
CoasterDataCleaned |>
  group_by(Type, Design) |>
  summarize(count = n()) |>
  pivot_wider(names_from = Type, values_from = count)
```

- Create a stacked bar graph for Type by InversionsYN

```{r}
ggplot(data = CoasterDataCleaned, aes(x = Type, fill = InversionsYN)) +
  geom_bar()+
  labs(x = "Type") +
  scale_fill_discrete("Inversions? Yes/No")
```

- Recreate the above as a side-by-side bar graph

```{r}
ggplot(data = CoasterDataCleaned, aes(x = Type, fill = InversionsYN)) +
  geom_bar(position = "dodge")+
  labs(x = "Type") +
  scale_fill_discrete("Inversions? Yes/No")
```

### Numeric Variables

- Find measures of center and spread for two numeric variables:  TopSpeed and MaxHeight

```{r}
CoasterDataCleaned |>
  summarize(across(.cols = c(TopSpeed, MaxHeight), 
            list("mean" = ~ mean(.x, na.rm = TRUE), "sd" = ~ sd(.x, na.rm = TRUE)),
            .names = "{.fn}_{.col}"))
```

- Now find center and spread of TopSpeed and MaxHeight for Extreme Scale roller coasters

```{r}
CoasterDataCleaned |>
  filter(Scale == "Extreme") |>
  summarize(across(.cols = c(TopSpeed, MaxHeight), 
            list("mean" = ~ mean(.x, na.rm = TRUE), "sd" = ~ sd(.x, na.rm = TRUE)),
            .names = "{.fn}_{.col}"))
```

Note:  Compared to all roller coasters, Extreme roller coasters have higher mean TopSpeed (56.9 vs 42.2) and higher mean MaxHeight (126.5 vs 76.8).

- Find measures of center and spread of Length and Duration across AgeGroup


```{r}
CoasterDataCleaned |>
  group_by(AgeGroup) |>
  summarize(across(.cols = c(Length, Duration), 
            list("mean" = ~ mean(.x, na.rm = TRUE), "sd" = ~ sd(.x, na.rm = TRUE)),
            .names = "{.fn}_{.col}"))
```


- Find measures of center and spread of Drop and NumInversions across Type and CensusRegion 

```{r}
CoasterDataCleaned |>
  group_by(Type, CensusRegion) |>
  summarize(across(.cols = c(Drop, NumInversions), 
            list("mean" = ~ mean(.x, na.rm = TRUE), "sd" = ~ sd(.x, na.rm = TRUE)),
            .names = "{.fn}_{.col}"))
```

- Create a correlation matrix between all of the numeric variables

```{r}
NumericVars <- CoasterDataCleaned |>
  select(where(is.numeric)) 

cor(NumericVars, use = "pairwise.complete.obs")
```

Note:  The three variables TopSpeed, MaxHeight, and Drop all have very strong positive correlations with each other (0.89 or greater).  Length also shows positive correlation with each of these variables (range 0.67 to 0.74).  

- Create a histogram for MaxHeight and TopSpeed across Type

```{r}
g <- ggplot(CoasterDataCleaned |> drop_na(MaxHeight), aes(x = MaxHeight))
g + geom_histogram(binwidth = 50, alpha = .5, aes(fill = Type))
g <- ggplot(CoasterDataCleaned |> drop_na(TopSpeed), aes(x = TopSpeed))
g + geom_histogram(binwidth = 15, alpha = .5, aes(fill = Type))
```

- Create a kernel density plot for MaxHeight and TopSpeed across Type

```{r}
g <- ggplot(CoasterDataCleaned |> drop_na(MaxHeight), aes(x = MaxHeight))
g + geom_density(alpha = .5, aes(fill = Type))
g <- ggplot(CoasterDataCleaned |> drop_na(TopSpeed), aes(x = TopSpeed))
g + geom_density(alpha = .5, aes(fill = Type))
```

- Create a boxplot for MaxHeight and TopSpeed across Type

```{r}
g <- ggplot(CoasterDataCleaned |> drop_na(MaxHeight))
g + geom_boxplot(aes(x = Type, y = MaxHeight, fill = Type))
g <- ggplot(CoasterDataCleaned |> drop_na(TopSpeed))
g + geom_boxplot(aes(x = Type, y = TopSpeed, fill = Type))
```

