---
title: "HW #4"
author: "Ryan Friedman, Evan Whitfield, Ralph Church"
format: html
editor: visual
---

```{r}
#| include: false
# load packages that may be used
library(tidyverse)
library(readxl)
library(DBI)
library(knitr)
library(httr)
library(jsonlite)
```

## Task 1 - Querying an API and a Tidy-Style Function

1.  Use GET() from the httr package to return information about a topic that you are interested in that has been in the news lately (store the result as an R object).

```{r}
RFKey <- "27554db1fa15488a995b7935d17bdf94" # R.Friedman API key for newsapi.org

APIData <- GET(paste0("https://newsapi.org/v2/everything?q=SpaceX&from=2026-06-01&language=en&pageSize=50&apiKey=", RFKey)) # Gets data on topic "SpaceX" from 6/1/2026 onward.
```

2.  Parse what is returned and find your way to the data frame that has the actual article information in it.

```{r}
parsed <- fromJSON(rawToChar(APIData$content))
parsed$articles
```

3.  Write a function that allows the user to easily query NewsAPI. The inputs are the title/subject to search for, a time period to search from (i.e. from that time until the present), and an API key.

```{r}
NewsAPIQuery <- function (subject,TimePeriod, NewsAPIKey){
  NewsURL <- paste0("https://newsapi.org/v2/everything?q=", subject,"&from=", TimePeriod, "&language=en&pageSize=50&apiKey=", NewsAPIKey) # creates the URL from the arguments
  NewsData <- GET(NewsURL) # gets the API data
  NewsParsed <- fromJSON(rawToChar(NewsData$content)) # convert content of API data into readable test
  NewsParsed$articles # returns articles dataframe
}
```

Testing function with search of "Disneyland" from "2026-06-08".

```{r}
NewsAPIQuery("Disneyland", "2026-06-08", RFKey)
```

Testing function with search of "Tokyo" from "2026-06-04".

```{r}
NewsAPIQuery("Tokyo", "2026-06-04", RFKey)
```

## Task 2 - Summarizing Roller Coaster Data

- Read in the data. Confirm variable data types and identify factors

```{r}
CoasterDataRaw <-read_csv("HW4DataFiles/635USrollercoasters.csv")
CoasterDataRaw
```

- Delete Latitude, Longitude, and Boundary variables (not needed for the assignment)
- Fix variable names
- Variables to be converted to factors: CensusRegion, AgeGroup, Scale, Type, Design, InversionsYN

```{r}
CoasterDataCleaned <- CoasterDataRaw |>
  select(-Latitude, -Longitude, -Boundary) # Remove undesired variables

colnames(CoasterDataCleaned) = c("Coaster", # Set variable names
                                 "Park",
                                 "City",
                                 "State",
                                 "CensusRegion",
                                 "YearOpened",
                                 "AgeGroup",
                                 "TopSpeed",
                                 "MaxHeight",
                                 "Scale",
                                 "Type",
                                 "Design",
                                 "Drop",
                                 "Length",
                                 "Duration",
                                 "InversionsYN",
                                 "NumInversions",
                                 "Make")  

CoasterDataCleaned <- CoasterDataCleaned |>
  mutate(CensusRegion = factor(CensusRegion), # Convert select variables to factors
         AgeGroup = factor(AgeGroup),
         Scale = factor(Scale),
         Type = factor(Type),
         Design = factor(Design),
         InversionsYN = factor(InversionsYN))

str(CoasterDataCleaned) # Confirm remaining variables and types
```
